\mbox{Specifying Imperative Higher-Order Programs}

### 2009-10-06 papers/esop2010/main.tex
remarques de JC

\author{Johannes Kanig\inst{1,}\inst{2} \and Jean-Christophe
Fill\^atre\inst{2,}\inst{1}}
\institute{
INRIA Saclay - Île-de-France, ProVal, Orsay, F-91894 \and
 LRI, Univ Paris-Sud, CNRS, Orsay, F-91405
}
%TODO explain somewhere the correspondence store type <-> map, effect variable
%<-> domain, region <-> key
%TODO the constants dealing with maps

@@

\begin{abstract} We present a specification language suited for reasoning
  about effectful higher-order programs. The language's main features are
  higher-order logic and effect polymorphism, both of which enable a truly
  modular treatment of higher-order functions.  We also define a weakest
  precondition calculus for this language which generates proof obligations in
  higher-order logic.  To our knowledge, this is the first time that effect
  polymorphism is put to use in program verification.  A simple mechanism
  based on regions permits to restrict the analysis to alias-free programs,
  thus obtaining much simpler proof obligations. Several examples demonstrate
  the practicality of the system. A prototype implementation exists and has
  been used to prove the correctness of these and more examples.
\end{abstract}

@@

The purpose of the present paper is to introduce a simple weakest precondition
calculus for imperative higher-order programs. Higher-order functions are a
key feature, not only of purely functional languages like Haskell and the pure
fragment of ML, but also in combination with imperative features, like
references. However, reasoning about higher-order functions is difficult, and
especially so in the presence of effectful computations. Any function's
specification depends on its arguments. In a higher-order setting, arguments
may be functions that have their own specification. In consequence, to be
modular, the specification of a higher-order function can only be formulated
{\em depending on the specifications of its functional arguments}. In a
setting where side effects are possible, one also needs to formulate the
specification {\em depending on the side effects of the functional arguments}.
We claim that to be truly modular, the specification of a higher-order
function must be generic with respect to the specification {\em and} the
effect of its arguments.

In this paper, we first present a specification language combining features of
a programming language and logic. It uses an effect system and a region
calculus to track references. The effect system contains {\em effect
polymorphism}, thus dealing well with higher-order functions. The
specifications can be expressed in higher-order logic, to be able to state
properties of functional arguments. We show how two very simple modifications
to the system let us obtain alias-free programs, with the aim of obtaining
much simpler proof obligations.

We then proceed to define a weakest precondition calculus based on this
language, transforming an annotated program into a logic formula, which has to
be valid for the program to be correct. This calculus is very straight-forward
and syntax-directed.

Finally, we show the usefulness of the system by giving several examples of
annotated programs. All the programs have been proved correct using our
prototype implementation called \who.

\paragraph{Outline.}
Section~\ref{sec:genpic} gives an introduction to the described language as
well as an example. Sections~\ref{sec:syntax} and \ref{sec:wp} give the
details about the programming language and the weakest precondition calculus.
In section~\ref{sec:examples} we speak about the implementation and show some
examples. Finally, section~\ref{sec:related} discusses related work.

@@

Consider the following first-order program\footnote{Throughout the document,
we use \ocaml~Syntax for programs.} that accumulates the sum of the contents
of some array $a$ in some global reference $c$, using an integer $i$ as
a counter:
\begin{ocaml}
  let sum () = 
    for i = 0 to Array.length a - 1 do
      c := !c + a.(i)
\end{ocaml}
This simple  can be annotated in a standard Hoare logic (for
example the system of \cite{Filliatre00a}) as follows:
\begin{who}
  let sum () =
  {!!c = 0}
    for i = 0 to length !a - 1 do
    { !!c = {{{ $\sum_0^{i-1}$ get !!a i }}} }

      c := !c + get !a i

      {!!c = {{{ $\sum_0^{ {\tt length~!a} - 1}$ get !!a i }}} }
\end{who}

Here, assuming we model the array in the programming language by a reference
to a functional array, with functions {\tt length} and {\tt get} with the
obvious meaning. Also, in the logic we use a derefence operator {\tt !!}
instead of the usual {\tt !}. In this case, our program is a {\tt for} loop,
so it needs an {\em invariant annotation}, indicating which relation is
preserved by the loop body. The equalities that need to be proved are very
simple. Also, it is obvious that the program modifies exactly the references
$r$ and $c$, and this fact can be formally confirmed by an effect analysis.

A common reaction of a functional programmer to a program like
$\mathit{sum}$ is to factor out the structure of the recursion, which
may be useful in other situations, in a higher-order function. The
more specific program is then recovered by function application. In this case,
the more general function is of course simply {\tt Array.iter}:
\begin{ocaml}
  let iter f a = 
    for i = 0 to Array.length a - 1 do
      f a.(i)

  let sum () =
    iter (fun n -> c := !c + n) a
\end{ocaml}
It is obvious that the two programs do exactly the same
thing. However, how do we prove the correctness of the second one? One
possibility is to fix the future usage of ${\tt iter}$ by
specifying what exactly is the specification of its argument {\tt f}:
\begin{who}
  let iter 
    (f : (x : int) ->{c} 
        { !!c = {{{ $\sum_0^{i-1}$ get !!a i }}} } 
        unit 
        { !!c = {{{ $\sum_0^{length !a - 1}$ get !!a i }}} } ) 
    (a : int array) =
    ...
\end{who}
Here we impose precisely the specification of {\tt f}, and thus we can
derive that {\tt sum} computes the right value,
using the same method as for the first-order case before. Note
that, to be able to apply an effect analysis, we have to specify that
$f$ modifies $c$, by giving it the type $\intml->^c\unitml$.

Of course, this approach is not very satisfactory. While the program achieves
the separation of the iterating function ({\tt iter}) and the combining
function (the anonymous function), the specification does not do this at all;
as specified,  {\tt iter} can basically only be applied to the same function,
and can only compute a certain sum. If one wants to use {\tt iter} in another
context, one has to copy the code, give a new specification adapted to the
application and prove this new specification again. In addition, in order to
be able to write the specification at all, we had to specialize the type of
{\tt iter} to integer arrays, while its \ocaml\ counterpart is of course
polymorphic. Said otherwise, this approach is not modular at all.

There are two obstacles related to higher-order functions to be
overcome in order to achieve modularity on the level of specification
in the presence of side effects. We must be able to refer to the
effect of a function parameter, and we must be able to refer to the
specification (pre- and postconditions) of a function parameter, in a
generic way.

The first obstacle has a well known solution using a type and effect
system~\cite{talpin94taed} with effect variables: effect polymorphism
does exactly what we want here. Using effect polymorphism, we can say
that {\tt f} (the argument of {\tt iter}) has type
$\intml->^e\unitml$, where $e$ is an {\em effect variable}, which
amounts to say that {\tt f} has a {\em certain} effect $e$. The total
effect of {\tt iter} is then $e$, meaning {\tt iter}
modifies whatever {\tt f} modifies.

@@

The main idea of this paper is to combine the effect-based Hoare logic of
\cite{Filliatre00a} with the more powerful effect system of
\cite{talpin94taed} and the more powerful annotations of
\cite{regis-gianas-pottier-08} to obtain a modular system for proving
higher-order programs with side effects. As in~\cite{regis-gianas-pottier-08},
we use operators \preml\ and \postml\ to access the pre- and postcondition of
a function. However, in our system, a precondition takes as an argument not
only the function argument, but also the initial state, before the execution
of the function. The postcondition takes the argument and return value, but
also the states before {\em and} after the function call. Th so-called
two-state postconditions are a way to get rid of most uses of auxiliary
variables. However, the pre- and postcondition of {\tt f} may not refer to
arbitrary parts of the state, but only the part {\tt f} reads or modifies.
With these conventions, one can now write a generic specification for the {\tt
iter} function, see ~Fig.~\ref{fig:iter}.

@@

There are several new elements here. First of all, we use effect polymorphism
to introduce an effect variable {\tt e} standing for the effect of {\tt f}.
Also, {\tt iter} has an additional argument {\tt inv}, the {\em invariant},
taking as argument the portion of the state modified by f and the integer
indicating the current iteration count. This is a logical argument, which
means that it cannot be used in the program. Also, {\tt inv} is a {\em pure}
function, which is indicated by the arrows without effect annotation. The
postcondition is now quite clear: we simply want to establish the invariant
for the final state (the keyword {\tt cur} for the current state of the
postcondition) and the length of the array, indicating that we have iterated
over the entire array. It is also easy to see that the invariant {\tt inv} is
precisely the invariant of the {\tt for} loop. However, the precondition is a
bit complicated: it states that we need the invariant to hold initially (for
the current state of the precondition and index $0$) and we need the invariant
to be preserved by the function call. This is expressed by quantifying over
two states, saying that the invariant has to imply the precondition of {\tt f}
and the postcondition has to imply the invariant for the next array index.
With a bit of syntactic sugar, this could be expressed much more concisely
using a Hoare triple:
\begin{who}
  { inv cur i} f (get i !!a) {inv cur (i+1)}
\end{who}

The {\tt iter} function and its specification raise a question: is the
specification correct if {\tt f} modifies the array {\tt a}, which is required
for the good behavior of {\tt iter}? The answer is clearly no; if {\tt f}
messes up the contents of {\tt a}, {\tt iter} may not actually maintain the
invariant.  This is an instance of the {\em aliasing problem} (see, for
example, \cite{reynolds78popl}) and it is well known that standard Hoare logic
does not support aliasing. In this paper we make the same choice and disallow
aliasing in the programming language we introduce. Under this assumption, the
specification we gave is correct.

Entirely excluding aliasing renders our system inappropriate for
reasoning about shared mutable structures. However, it is interesting
to see that our alias exclusion restriction, in many cases,
corresponds to a natural condition imposed on higher-order functions
that operate on some data structure. It is a common
requirement~\cite{Nanevski08Awkward,Krishnaswami06} for iteration
functions, for example, that the functional argument must not modify
the structure to be iterated on. In our system, this fact follows from
the restriction on effect instantiations, as will be seen in
section~\ref{sec:syntax}. In the example, the array {\tt a}
of the iterating function {\tt iter} must not be modified by its
argument.

@@

Before we describe the syntax of types, let us first explain {\em regions} and
{\em effects}. Regions are simply type level names for memory locations. An
effect is a set of memory locations and {\em effect variables}. An effect
variable stands for an unknown effect. Effects are used to specify
of memory locations an expression can write. They are also used in function
types to describe the memory locations which may be written if the function is
called.

@@

\paragraph{Value typing.}
Let us first look at the common part concerning values. We define a relation
$Γ|-_v t : τ$, which means ``$t$ is well-typed wrt. the environment $Γ$, and
is of type $τ$''. Values cannot have any effect.  Variables and pure
abstractions are typed in the usual way; type instantiations at variable
occurrences are substituted in the type. The rule {\sc REC} simply states that
the body of the function can use the argument $x$ as well as the function $f$.
The specifications $p$ and $q$, however, cannot use $f$ and must be well-typed
wrt. the typing relation $|-_l$ described later on. Finally, pure applications
between values are considered as values as well, as they don't have any
effect.

@@

\paragraph{Program typing.}
The typing of programs, written $Γ|- t : τ,ε$, assigns a type $τ$ and an
effect $ε$ to an expression $t$, given an environment. This typing is
presented in {\em A-normal form}, a normal form where all intermediate values
are \letml-introduced. This is not a restriction, as any program can be
transformed an A-normal form.

All values are also programs with no effect (rule {\sc
Value}). An application must consist of two values and the effect of the
overall expression is precisely the effect of the called function (rule {\sc
App}). There are two rules for \letml, due to the value restriction. One only
can generalize type, effect and region variables when the bound expression is
a value. In this case (rule {\sc LetV}), only the second term may produce
effects. In the other case (rule {\sc Let}), no generalization takes place,
but both terms can have effects. The {\sc Let} rule actually obliges both
terms to have the same effect, but together with the {\sc Sub} rule, which can
increase the effect of an expression, this can always be achieved.
Finally, \ifml-expressions are typed in the usual way, both branches have to
have the same effect.

@@

\begin{figure}[tbp]
  \begin{mathpar}
  { \inferrule*[Left=Logic]
    { Γ |-_v t : τ  }
    { Γ |-_l t : \ceil{τ} }
  } \and
  { \inferrule*[Left=LetLogic]
    { Γ, \alist{χ} |-_l : t' : τ' \\ Γ, x: ∀\alist{χ}.τ' |-_l t : τ  }
    { Γ |-_l \letml~x~[\alist{χ}] = t'~\inml~t : τ }
  }\and
  { \inferrule*[Left=IteLogic]
    { Γ |- _l t1 : \boolml \\ Γ |-_l t2 : τ \\ Γ |-_l t3 : τ }
    { Γ |- \ifml~ t1 ~\thenml~ t2 ~\elseml ~ t3 : τ }
  }
  \end{mathpar}
  \caption{Logic typing}
  \label{fig:logictyping}
\end{figure}

@@

The reader may have wondered why the specification language does not contain
any logical connectives. These can of course be declared independently, for
example the conjunction is just a function: \begin{equation*} /\ :  \propml ->
\propml -> \propml \end{equation*} Actually, when equality is given, all
logical operations (even quantification) can be defined in the
logic~\cite{andrews86}.  In the remainder of the paper, we suppose
predefined the usual logical connectives conjunction, disjunction,
implication and equivalence, as well as universal and existential
quantification. We also assume all the arithmetic operations as well as the
constructor $(t,t)$  and the accessors $fst$ and $snd$ for pairs. We will also
use the names \preml\ and \postml\ for $fst$ and $snd$, respectively.

@@

We haven't mentioned references yet, but all we need to support them is to add
the three usual functions $ref, :=, !$ for reference creation, assignment and
lookup, respectively, to the environment. In our system, these functions have
the following types:
\begin{eqnarray*}
  ref &:& ∀αρ.α ->^ρ α~\refml_ρ\\
  := &:& ∀αρ.α~\refml_ρ ->^\emptyset α ->^ρ \unitml\\
  ! &:& ∀αρ. α~\refml_ρ ->^ρ α
\end{eqnarray*}

@@

A way out is to restrict programs to alias free programs. Fortunately, it is
easy to modify the presented typing rules to render all programs with aliasing
ill-typed (see Fig~\ref{fig:aliasing}). In the following, we will use regions
to identify memory locations entirely; each region shall describe precisely
one memory location. Starting from this assumption, a program can actually
exhibit {\em two} forms of aliasing. In the first one, it may use several
names (region variables) for the same memory location. This can happen when a
function is generalized over, say, two region variables, assuming they are
different, and is then instantiated with the same region variable twice. The
classical example is the following program
\begin{who}
  let f [ρ ρ'] (x : int ref(ρ)) (y : int ref(ρ')) =
  { }
  x := !x + 1;
  !y
  { r : r = !!y old }
\end{who}
which is only correct if $ ρ $ and $ ρ' $ are indeed different. To enforce
this, we add a side condition to the instantiation: the substitution $σ$ must
be {\em compatible} with the type $τ$. To understand what compatibility means,
consider an {\em atomic} region substitution (whose domain contains only one
variable) $[ρ|->ρ']$. It is compatible with any type which does not contain
$ρ'$. Similarly, the effect substitution $[e|-> ρ1  ρ2 e'$] is compatible with
any type which does not contain $ ρ1 , ρ2 $ nor $e'$.  Expressed more
generally, an atomic substitution is compatible with a type if the type does
not contain any region or effect variables which are in the range of the
substitution. Finally, a {\em composite} substitution (whose domain contains more
than one variable) is compatible
with a type if each atomic substitution is compatible with the result of the
application of the previous substitution. In this way, it becomes impossible
to have different names for a single reference cell.

The second form of aliasing is when we use the same name for {\em different}
memory locations. This can happen when we allow the user to call $ref$ several
times with the same region instantiation. A simple possibility is to attach
the suppress the $ref$ function entirely and to replace it by a \letml\refml\ 
construct which combines name creation and memory allocation. In this way,
there can of course be only one reference per name. On the other hand, because
of the side condition $ρ\notin τ$, the reference cannot escape the scope of
its \letml\refml\ construct, and so, a function cannot return a reference
which has been created in its body. This restriction could be leveraged a bit if
one separates name creation and allocation, while still maintaining that one
can only allocate a single reference in each region. This is how it is done in
the correctness proof of this restriction. The presented \letml\refml\
construct is of course a special case.
%TODO now, all operations on maps can be statically resolved.

@@

Subject reduction of our type system is a consequence of subject reduction of
the region calculus of Tofte/Talpin~\cite{tofte97ic}.

\section{A Simple Weakest Preconditions Calculus}
\label{sec:wp}

The aim of the weakest precondition calculus is to obtain proof obligations
from the annotated program. To do this, the calculus starts from a formula
which holds, or is supposed to hold, at the end of the program, and walks
backwards in the program source to obtain the weakest formula which has to
hold at the beginning of the program (hence the name {\em weakest
precondition}.

@@

As for the typing relation, we have two rules for the \letml-binding. If the
bound term is a value, we can translate the \letml-binding in the program by a
\letml-binding in the language. In the other case, the \wpre\ of $ t2 $ becomes
the argument of the \wpre of $ t1 $ (we denote $ε$ the effect of $ t1 $ and $
t2 $). The rule for \ifml-expressions is straightforward. The rule for
subeffecting adapts the domain of the current store and the expected domain of
the postcondition so that the types work out.

>>>>

\author{Johannes Kanig\inst{1,}\inst{2} \and Jean-Christophe
Fill\^atre\inst{2,}\inst{1}}
\institute{
INRIA Saclay - Île-de-France, ProVal, Orsay, F-91893 \and
 LRI, Univ Paris-Sud, CNRS, Orsay, F-91405
}
%TODO explain somewhere the correspondence store type <-> map, effect variable
%<-> domain, region <-> key
%TODO the constants dealing with maps

@@

\begin{abstract} We present a specification language suited for reasoning
  about effectful higher-order programs. The language's main features are
  higher-order logic and effect polymorphism, both of which enable a truly
  modular treatment of higher-order functions.  We also define a weakest
  precondition calculus for this language which generates proof obligations in
  higher-order logic.  To our knowledge, this is the first time that effect
  polymorphism is put to use in program verification. Additionally, we propose
  a region-based analysis to exclude aliasing, in order to obtain much simpler
  proof obligations. Several examples demonstrate the practicality of the
  system. A prototype implementation exists and has been used to prove the
  correctness of these and more examples.  \end{abstract}

@@

The purpose of the present paper is to propose a language suited for reasoning
about effectful higher-order programs.  Higher-order functions are a key
feature, not only of purely functional languages like Haskell and the pure
fragment of ML, but also in combination with imperative features, like
references or arrays. However, reasoning about higher-order functions is
difficult, and especially so in the presence of effectful computations. Any
function's specification depends on its arguments. In a higher-order setting,
arguments may be functions that have their own specification. In consequence,
to be modular, the specification of a higher-order function can only be
formulated {\em depending on the specifications of its functional arguments}.
In a setting where side effects are possible, one also needs to formulate the
specification {\em depending on the side effects of the functional arguments}.
We claim that to be truly modular, the specification of a higher-order
function must be generic with respect to the specification {\em and} the
effect of its arguments.

In this paper, we first present a specification language combining features of
a programming language and logic. It uses an effect system and a region
calculus to track references. The effect system contains {\em effect
polymorphism}, thus dealing well with higher-order functions. The
specifications can be expressed in higher-order logic, to be able to state
properties of functional arguments.  We then proceed to define a weakest
precondition calculus for this language, transforming an annotated
program into a logic formula, which has to be valid for the program to be
correct. This calculus is very straight-forward and syntax-directed.

Programs in our language may contain aliases. However, we obtain much simpler
proof obligations when aliases are excluded. Therefore, we also propose a
static region-based analysis rejecting programs with aliasing. The weakest
precondition calculus on the alias-free fragment has been implemented in a
prototype tool called \who~\cite{KanigFilliatre09wml}.  We show the usefulness
of this tool on several examples, which have all been proved correct.

\paragraph{Outline.}
Section~\ref{sec:genpic} introduces our language using a motivating example.
Sections~\ref{sec:syntax} and \ref{sec:wp} give the details about the
programming language and the weakest precondition calculus.  In
section~\ref{sec:examples} we present our implementation and show some
examples. Finally, section~\ref{sec:related} discusses related work.

@@

Consider the following first-order program\footnote{Throughout the document,
we use \ocaml~syntax for programs.} that accumulates the sum of the contents
of some array {\tt a} in some global reference {\tt c}, using an integer {\tt
i} as a counter:
\begin{ocaml}
  let sum () = 
    for i = 0 to Array.length a - 1 do
      c := !c + a.(i)
    done
\end{ocaml}
This simple program can be annotated in a standard Hoare logic (for
example the system of \cite{Filliatre00a}) as follows:
\begin{who}
  let sum () =
    { !c = 0 }
    for i = 0 to Array.length a - 1 do
      { !c = {{{ $\sum_0^{i-1}$ }}}  a.(i) }
      c := !c + a.(i)
    done
    { !c = {{{ $\sum_0^{ {\tt Array.length~a} - 1}$ }}}  a.(i) }
\end{who}
Our program is a {\tt for} loop, so it requires an {\em invariant
annotation}, indicating which relation is preserved by the loop body. The
equalities that need to be proved are very simple. Also, it is obvious that
the program only modifies the reference {\tt c}, and this fact can be
formally confirmed by an effect analysis.

A common reaction of a functional programmer to a program like
{\tt sum} is to factor out the structure of the recursion, which
may be useful in other situations, in a higher-order function. In this case,
the more general function is the following iterator over arrays:
\begin{ocaml}
  let iter f a = 
    for i = 0 to Array.length a - 1 do
      f a.(i)
    done
\end{ocaml}
The more specific program is then recovered by function application. 
\begin{ocaml}
  let sum () =
    iter (fun n -> c := !c + n) a
\end{ocaml}
It is obvious that the two programs do exactly the same
thing. However, how do we prove the correctness of the second one? One
possibility is to impose the future usage of ${\tt iter}$ by
specifying a particular specification for its argument {\tt f}:
\begin{who}
  let iter 
    (f : (x : int) ->{c} 
        { !c = {{{ $\sum_0^{i-1}$ }}} a.(i) } 
        unit 
        { !c = {{{ $\sum_0^{ {\tt Array.length~a} - 1}$ }}} a.(i) } ) 
    (a : int array) =
    ...
\end{who}
Here we impose precisely the specification of {\tt f}, and thus we can
derive that {\tt sum} computes the expected value,
using the same method as for the first-order case before. Note
that, to be able to apply an effect analysis, we have to specify that
$f$ modifies $c$, by giving it the type $\intml->^{\{c\}}\unitml$.

Of course, this approach is not very satisfactory. While the program achieves
the separation of the iterating function ({\tt iter}) and the combining
function (the anonymous function), the specification does not do this at all;
as specified,  {\tt iter} can basically only be applied to the same function,
and can only compute a certain sum. If one wants to use {\tt iter} in another
context, one has to copy the code, give it a new specification adapted to the
application and prove this new specification again. In addition, in order to
be able to write the specification at all, we had to specialize the type of
{\tt iter} to integer arrays, while its \ocaml\ counterpart is of course
polymorphic. Said otherwise, this approach is not modular at all.

There are two obstacles related to higher-order functions to be overcome in
order to achieve modularity on the level of specification in the presence of
side effects. We must be able to refer to the effect of a function parameter,
and we must be able to refer to its specification (pre- and postconditions),
in a generic way.

The first obstacle has a well known solution using a type and effect
system~\cite{talpin94taed} with effect variables: effect polymorphism does
exactly what we want here. Using effect polymorphism, we can say that {\tt f}
(the argument of {\tt iter}) has type $α->^e\unitml$, whera $α$ is the type of
the array elements and where $e$ is an {\em effect variable}, which amounts to
say that {\tt f} has a {\em certain} effect $e$. The complete type of {\tt
iter} is thus:
\begin{displaymath}
  ∀α∀e. (α ->^e \unitml) ->^\emptyset α~\mathtt{array} ->^e \unitml
\end{displaymath}
The total effect of {\tt iter} is then $e$, meaning {\tt iter}
modifies whatever {\tt f} modifies.

@@

The main idea of this paper is to combine the effect-based Hoare logic of
\cite{Filliatre00a} with the more powerful effect system of
\cite{talpin94taed} and the more powerful annotations of
\cite{regis-gianas-pottier-08} to obtain a modular system for proving
higher-order programs with side effects. As in~\cite{regis-gianas-pottier-08},
we use operators \preml\ and \postml\ to access the pre- and postcondition of
a function. However, in our system, a precondition takes as an argument not
only the function argument, but also the initial state, before the execution
of the function. The postcondition takes the argument and return value, but
also the states before {\em and} after the function call. Two-state
postconditions are a standard way to get rid of most auxiliary variables.
However, the pre- and postcondition of {\tt f} may not refer to arbitrary
parts of the state, but only the part {\tt f} reads or modifies.  With these
conventions, one can now write a generic implementation and specification for
the {\tt iter} function. It is given~Fig.~\ref{fig:iter}, in the input syntax
of our prototype \who. There are a few syntactic differences with \ocaml, in
particular for arrays.

%Here, assuming we model the array in the programming language by a reference
%to a functional array, with functions {\tt length} and {\tt get} with the
%obvious meaning. Also, in the logic we use a derefence operator {\tt !!}
%instead of the usual {\tt !}.

@@

There are several new elements here. First of all, we use effect polymorphism.
An effect variable {\tt e}, standing for the effect of {\tt f}, is introduced
by the notation ${\mathtt [e]}$.  {\tt iter} also has an additional argument
{\tt inv}, the {\em invariant}, taking as argument the portion of the state
modified by {\tt f} and the integer indicating the current iteration count.
This is a logical argument, which means that it cannot be used in the program.
Also, {\tt inv} is a {\em pure} function, which is indicated by the arrows
without effect annotation. The postcondition is now quite clear: we simply
want to establish the invariant for the final state (the keyword {\tt cur}
stands for the current state of the postcondition) and the length of the
array, indicating that we have iterated over the entire array. It is also easy
to see that the invariant {\tt inv} is precisely the invariant of the {\tt
for} loop.  However, the precondition is a bit complicated: it states that we
need the invariant to hold initially (for the current state of the
precondition and index $0$) and we need the invariant to be preserved by the
function call.  This is expressed by quantifying over two states, saying that
the invariant has to imply the precondition of {\tt f} and the postcondition
has to imply the invariant for the next array index.  With a bit of syntactic
sugar, this could be expressed much more concisely using a Hoare triple:
\begin{who} 
  { inv cur i } f (get i !!a) { inv cur (i+1) } 
\end{who}

% TODO rewrite
%The {\tt iter} function and its specification raise a question: is the
%specification correct if {\tt f} modifies the array {\tt a}, which is required
%for the good behavior of {\tt iter}? The answer is clearly no; if {\tt f}
%messes up the contents of {\tt a}, {\tt iter} may not actually maintain the
%invariant.  This is an instance of the {\em aliasing problem} (see, for
%example, \cite{reynolds78popl}) and it is well known that standard Hoare logic
%does not support aliasing. In this paper we make the same choice and disallow
%aliasing in the programming language we introduce. Under this assumption, the
%specification we gave is correct. 
%
%Entirely excluding aliasing renders our system inappropriate for
%reasoning about shared mutable structures. However, it is interesting
%to see that our alias exclusion restriction, in many cases,
%corresponds to a natural condition imposed on higher-order functions
%that operate on some data structure. It is a common
%requirement~\cite{Nanevski08Awkward,Krishnaswami06} for iteration
%functions, for example, that the functional argument must not modify
%the structure to be iterated on. In our system, this fact follows from
%the restriction on effect instantiations, as will be seen in
%section~\ref{sec:syntax}. In the example, the array {\tt a}
%of the iterating function {\tt iter} must not be modified by its
%argument.

@@

Before we describe the syntax of types, let us first explain {\em regions} and
{\em effects}. Regions are simply type level names for memory locations. An
effect is a set of memory locations and {\em effect variables}. An effect
variable stands for an unknown effect. Effects are used to specify
the memory locations an expression can write. They are also used in function
types to describe the memory locations which may be written if the function is
called.

@@

\paragraph{Value typing.}
Let us first look at the common part concerning values. We define a relation
$Γ|-_v t : τ$, which means ``$t$ is well-typed wrt. the environment $Γ$, and
is of type $τ$''. Values cannot have any effect.  Variables and pure
abstractions are typed in the usual way; type instantiations at variable
occurrences are substituted in the type. The rule {\sc REC} simply states that
the body of the function can use the argument $x$ as well as the function $f$.
The specifications $p$ and $q$, however, cannot use $f$ and must be well-typed
wrt. the typing relation $|-_l$ described later on. Finally, pure applications
between values are considered to be values as well, as they don't have any
effect.

@@

\paragraph{Program typing.}
The typing of programs, written $Γ|- t : τ,ε$, assigns a type $τ$ and an
effect $ε$ to an expression $t$, given an environment. This typing is
presented in {\em A-normal form}, a normal form where all intermediate values
are \letml-introduced. This is not a restriction, as any program can be
transformed into A-normal form.

All values are also programs with no effect (rule {\sc
Value}). An application must consist of two values and the effect of the
overall expression is precisely the effect of the called function (rule {\sc
App}). There are two rules for \letml, due to the value restriction. One only
can generalize type, effect and region variables when the bound expression is
a value. In this case (rule {\sc LetV}), only the second term may produce
effects. In the other case (rule {\sc Let}), no generalization takes place,
but both terms can have effects. The {\sc Let} rule actually requires both
terms to have the same effect, but together with the {\sc Sub} rule, which can
increase the effect of an expression, this can always be achieved.
Finally, \ifml-expressions are typed in the usual way, both branches have to
have the same effect.

@@

\begin{figure}[tbp]
  \begin{mathpar}
  { \inferrule*[Left=Logic]
    { Γ |-_v t : τ  }
    { Γ |-_l t : \ceil{τ} }
  } \qquad \and
  { \inferrule*[Left=LetLogic]
    { Γ, \alist{χ} |-_l : t' : τ' \\ Γ, x: ∀\alist{χ}.τ' |-_l t : τ  }
    { Γ |-_l \letml~x~[\alist{χ}] = t'~\inml~t : τ }
  }\and
  { \inferrule*[Left=IteLogic]
    { Γ |- _l t1 : \boolml \\ Γ |-_l t2 : τ \\ Γ |-_l t3 : τ }
    { Γ |- \ifml~ t1 ~\thenml~ t2 ~\elseml ~ t3 : τ }
  }
  \end{mathpar}
  \caption{Logic typing}
  \label{fig:logictyping}
\end{figure}

@@

The reader may have wondered why the specification language does not contain
any logical connectives. These can of course be declared independently, for
example the conjunction is just a function: 
\begin{equation*} 
  /\ :  \propml -> \propml -> \propml 
\end{equation*} 
Actually, when equality is given, all logical operations (even quantification)
can be defined in the logic~\cite{andrews86}.  In the remainder of the paper,
we suppose predefined the usual logical connectives conjunction, disjunction,
implication and equivalence, as well as universal and existential
quantification. We also assume all the arithmetic operations as well as the
constructor $(t,t)$  and the accessors $\mathit{fst}$ and $\mathit{snd}$ for
pairs. We will also use the names \preml\ and \postml\ for $\mathit{fst}$ and
$\mathit{snd}$, respectively.

@@

We haven't mentioned references yet, but all we need to support them is to add
the three usual functions $\mathit{ref}, :=, !$ for reference creation, assignment and
lookup, respectively, to the environment. In our system, these functions have
the following types:
\begin{eqnarray*}
  \mathit{ref} &:& ∀αρ.α ->^ρ α~\refml_ρ\\
  := &:& ∀αρ.α~\refml_ρ ->^\emptyset α ->^ρ \unitml\\
  ! &:& ∀αρ. α~\refml_ρ ->^ρ α
\end{eqnarray*}

@@

A way out is to restrict programs to alias free programs. Fortunately, it is
easy to modify the typing rules to render all programs with aliasing
ill-typed (see Fig~\ref{fig:aliasing}). In the following, we will use regions
in such a way that there is exactly one memory location in each region.
Starting from this assumption, a program can actually
exhibit {\em two} forms of aliasing. In the first one, it may use several
names (region variables) for the same memory location. This can happen when a
function is generalized over, say, two region variables, assuming they are
different, and is then instantiated with the same region variable twice. The
classical example is the following program
\begin{who}
  let f [ρ ρ'] (x : int ref(ρ)) (y : int ref(ρ')) =
  { }
  x := !x + 1;
  !y
  { r : r = !!y old }
\end{who}
which is only correct if $ ρ $ and $ ρ' $ are indeed different. To enforce
this, we add a side condition to the instantiation: the substitution $σ$ must
be {\em compatible} with the type $τ$, written $σ\sim τ$. To understand what
compatibility means, consider an {\em atomic} region substitution (whose
domain contains only one variable) $[ρ|->ρ']$. It is compatible with any type
which does not contain $ρ'$. Similarly, the effect substitution $[e|-> ρ1  ρ2
e'$] is compatible with any type which does not contain $ ρ1 , ρ2 $ nor $e'$.
Expressed more generally, an atomic substitution is compatible with a type if
the type does not contain any region or effect variables which are in the
range of the substitution. Finally, a {\em composite} substitution (whose
domain contains more than one variable) is compatible with a type if each
atomic substitution is compatible with the result of the application of the
previous substitution. In this way, it becomes impossible to have different
names for a single reference cell.

The second form of aliasing is when we use the same name for {\em different}
memory locations. This can happen when we allow the user to call
$\mathit{ref}$ several times with the same region instantiation. A simple
possibility is to  suppress the $\mathit{ref}$ function entirely and to
replace it by a \letml\refml\ construct which combines name creation and
memory allocation. In this way, there can of course be only one reference per
name. On the other hand, because of the side condition $ρ\notin τ$, the
reference cannot escape the scope of its \letml\refml\ construct, and so, a
function cannot return a reference which has been created in its body. This
restriction could be leveraged a bit if one separates name creation and
allocation, while still maintaining that one can only allocate a single
reference in each region. This is how it is done in the correctness proof of
this restriction. The presented \letml\refml\ construct is of course a special
case.
%TODO now, all operations on maps can be statically resolved.

@@

Subject reduction of our type system is a consequence of subject reduction of
the region calculus of Tofte/Talpin~\cite{tofte97ic}.
%TODO ref vers l'annexe -  remarque sur la difference

\section{A Weakest Preconditions Calculus}
\label{sec:wp}

The aim of the weakest precondition calculus is to obtain proof obligations
from annotated programs. To do this, the calculus starts from a formula
which holds, or is supposed to hold, at the end of the program, and walks
backwards in the program source to obtain the weakest formula which has to
hold at the beginning of the program (hence the name {\em weakest
precondition}).

@@

As for the typing relation, we have two rules for the \letml-binding. If the
bound term is a value, we can translate the \letml-binding in the program by a
\letml-binding in the language. In the other case, the \wpre\ of $ t2 $ becomes
the argument of the \wpre\ of $ t1 $ (we denote $ε$ the effect of $ t1 $ and $
t2 $). The rule for \ifml-expressions is straightforward. The rule for
subeffecting adapts the domain of the current store and the expected domain of
the postcondition so that the types work out.
